An Imaging Search for Post-Main-Sequence Planets of Sirius B

### 2021-11-23 paper/paper.tex
remove commented text
gary/mike revisions

In recent decades thousands of exoplanets have been discovered orbiting stars that will eventually leave the stability of the main-sequence (MS) \citep{akesonNASAExoplanetArchive2013}. The eventual fate of planets around these stars is uncertain due to the large expansion, stellar winds, and high irradiation encountered during giant branch evolution \citep{verasPostmainsequencePlanetarySystem2016}. Despite this, direct evidence from white dwarf pollution \citep{juraExternallyPollutedWhite2007,xuSpitzerObservationsWhite2012}, debris disks \citep{deruyterKeplerianDiscsPostAGB2006,zuckermanAncientPlanetarySystems2010,koesterFrequencyPlanetaryDebris2014}, and substellar companions \citep[e.g.,][]{luhmanDiscoveryCandidateCoolest2011,vanderburgGiantPlanetCandidate2020,blackmanJovianAnalogueOrbiting2021} culminate to suggest planetary systems beyond the MS are more common than previously thought. Discovery and characterization of post-MS planets is essential to study planetary system evolution and planet-star interactions during the most critical phases of stellar evolution.

The evolution of intermediate-mass stars (\qtyrange{1}{8}{\solarmass}) comprises a violent and relatively brief period of giant branch evolution before all nuclear fusion ends and the stars become white dwarfs. As the main-sequence star runs out of hydrogen to burn in its core, it expands to 100s of times its size, engulfing any companions within the stellar radius. When helium fusion ignites, the giant star becomes three to four orders of magnitude brighter, causing stellar winds and strong irradiation. Eventually, the star runs out of fuel and concludes its nuclear burning, becoming a white dwarf. The white dwarf begins cooling, becoming three to four orders of magnitude dimmer than its MS progenitor.

@@

Previous searches for substellar companions around white dwarfs \citep[e.g.,][]{debesAreThereUnstable2002,hoganDODOSurveyII2009,luhmanDiscoveryCandidateCoolest2011,xuExtremeAOSearchGiant2015a} have primarily focused on detecting wide-orbit planets which survived the giant branch evolution of their hosts. Recent evidence and theoretical works, though, suggest ``forbidden'' regions of planetary evolution are worth investigating for exotic post-MS planets. These planets would provide crucial insight into planetary system evolution and planet-star interactions during giant branch evolution.

Direct imaging is well-suited for finding planets around white dwarfs due to the intrinsic faintness of the host (compared to a MS star) and lack of spectral features for radial velocity studies \citep{burleighImagingPlanetsNearby2002a}. With direct imaging, exoplanets can be detected and characterized independently from the orbital period of the planet, the stellar radius, or stellar spectrum, which makes it powerful compared to broadband photometric analysis or transit photometry methods. High-contrast imaging pushes direct imaging to its limits using large-aperture telescopes, adaptive optics, coronagraphy, low-noise detectors, image-processing, and observational techniques. Despite the power of high-contrast imaging, very few observations have been carried out for white dwarfs (a recent example is \citealt{pathakHighContrastImaging2021}).

Exoplanets are faint enough that only a few photons per second reach the detector. Host stars are typically many orders of magnitude brighter than the thermal emission of giant planets ($\sim10^{\text{-}8}$), and planets have very small angular separations from their hosts, which makes it difficult to disentangle the planet-light from the stellar diffraction pattern \citep{traubDirectImagingExoplanets2010}. In addition, atmospheric seeing and instrumental aberrations greatly reduce the sensitivity to exoplanets due to the manifestation of quasi-random perturbations of the instrumental point-spread function (speckles, see \citealp{guyonExtremeAdaptiveOptics2018}). Adaptive optics (AO) corrects a large percentage of the effects of speckles but has decreasing efficacy for dim targets due to the photon-limited nature of modern AO instrumentation. AO is also paramount for enabling coronagraphy, which attenuates on-axis starlight while transmitting off-axis signal, up to some inner working angle.

Typical high-contrast targets are nearby young objects observed in the near-infrared. Younger planets are brighter, thanks to their latent formation heat \citep{fortneyGiantPlanetInterior2010}, which reduces the star-planet contrast, and observing them in the near-infrared corresponds with their peak blackbody emission. Nearby targets have larger angular separations between potential planets and their hosts, which makes it easier to separate the signals. Nearby targets are also intrinsically brighter, enabling effective AO control. White dwarfs are atypical high-contrast targets due to their age and sparsity, but their relative faintness is compelling for the reduced contrast. In addition, a post-MS planet would be much younger than the system age of a white dwarf and would therefore still retain a large portion of its latent heat, further motivating high-contrast searches of nearby white dwarfs.

We set out in this work to perform high-contrast observations of Sirius B for post-MS planets. In the rest of this report, we will introduce the Sirius system as a potential candidate for post-MS planets, along with previous studies of white dwarf Sirius B (\Cref{sec:sirius}). We will detail our 2020 near-infrared observations of Sirius B with Keck/NIRC2, as well as our processing steps and statistical analysis for companion detection (\Crefrange{sec:obs}{sec:analysis}). Lastly, we will discuss our results within the context of Sirius and post-MS systems, as well as future directions for post-MS imaging (\Crefrange{sec:results}{sec:conclusion}).

% High-contrast imaging is a powerful, but challenging, technique for discovering and characterizing exoplanets. Giant exoplanets are most easily detected by the thermal emission of their latent formation heat, which peaks in the infrared and decreases as the planet ages \citep{fortneyGiantPlanetInterior2010}. The typical astrophysical flux ratios (contrasts) for a Sun-Jupiter analog in the near-infrared (NIR) are $\sim$$10^{\text{-}8}$, and for a Sun-Earth system are $\sim$$10^{\text{-}10}$ \citep{traubDirectImagingExoplanets2010}. The close angular separations of exoplanets from their host stars make it difficult to disentangle planet-light from the stellar diffraction pattern. This problem is exacerbated on the ground as the effects of atmospheric seeing further degrade the sensitivity to exoplanets. High-contrast imaging typically comprises large-aperture telescopes, adaptive optics (AO), coronagraphy, low-noise detectors, observational techniques, and image-processing to overcome the challenges in detecting the faint exoplanet signal.

% Using large-aperture telescopes provides greater collecting power and angular resolution. AO corrects, to an extent, the effects of atmospheric seeing and instrumental noise (speckles) by measuring wavefront errors and cancelling them out with dynamical optics, like a deformable mirror. Wavefront errors are measured from a wavefront sensor which typically is illuminated by the target itself. The high loop rates of AO, especially on the ground ($\sim$\qty{100}{\hertz} to \qty{1}{\kilo\hertz}), makes Poisson photon noise the limiting factor for AO, which limits the efficacy of faint imaging targets \citep{guyonExtremeAdaptiveOptics2018}. AO correction is necessary for maintaining the necessary angular resolution and sensitivity for exoplanet imaging, as well as keeping the stellar point-spread-function (PSF) centered on a coronagraph, if present.

% Coronagraphs are designed to control and attenuate the diffraction of starlight through the telescope while allowing off-axis exoplanet light to pass unperturbed. Coronagraphs are extremely sensitive to wavefront errors, particularly tip-tilt pointing errors, necessitating high-quality AO correction. Coronagraphs have an inner working angle, the separation where stellar throughput is halved, which is analogous to the full-width at half-maximum (FWHM) of a non-coronagraphic PSF. The inner working angle determines how close we can probe to the stars, and it fundamentally limits imaging searches to nearby systems, whose projected separations are closer to solar-system scales.

% Exoplanets are faint enough that only a few photons per second reach the detector \citep{traubDirectImagingExoplanets2010}, which makes high-gain, low-read-noise detectors paramount for direct imaging. Observational techniques, like angular differential imaging \citep{maroisAngularDifferentialImaging2006}, spatially decorrelate speckles to improve the noise attenuation of long observations. Post-processing techniques like PSF subtraction remove the stellar diffraction pattern, but at the risk of removing exoplanet signal. Altogether, these techniques have proven sufficient for ground-based telescopes to image giant planets outside our solar system.

% White dwarfs are interesting targets for direct imaging- they are three to four orders of magnitude fainter than their MS progenitors, reducing the contrast necessary to detect a companion. The lack of spectral features in white dwarfs \citep{schatzmanTheorieDebitEnergie1945} makes them difficult targets for the radial velocity technique. The small stellar radii of white dwarfs lowers the transit probability significantly, although detections of substellar companions around white dwarfs have been made \citep[e.g.,][]{vanderburgGiantPlanetCandidate2020}. The intrinsic faintness of white dwarfs proves challenging due to the high signal-to-noise ratio (S/N) required by AO. This is exacerbated on ground-based telescopes which require extreme AO to counteract the effects of atmospheric seeing. This constrains potential companion searches to relatively bright targets ($m^R\lesssim$ 13). Implicitly, this also limits searches to nearby white dwarfs ($d\sim$\qty{25}{\parsec}, \citealp{holberg25ParsecLocal2016}), which is beneficial for imaging because the limited inner working angle of the instrumentation will project to closer separations around closer stars.
\pagebreak

@@

One fascinating target for post-MS observations is Sirius B, the closest and brightest white dwarf to the sun. The Sirius system is the 7th closest to the sun at \qty{2.7}{\parsec}, consisting of Sirius A, a $J=$-1.36 magnitude A1V star, and Sirius B, a DA white dwarf with a \qty{50}{\year} orbit \citep{gaiacollaborationGaiaMission2016,bondSiriusSystemIts2017,gaiacollaborationGaiaEarlyData2021}. As mentioned previously, the proximity and intrinsic faintness of Sirius B (compared to a MS star) make it compelling for direct imaging. Additionally, the young system age ($\sim$\qty{225}{\mega\year}) means any giant planets would still retain much of their latent formation heat, increasing their luminosity in the IR \citep{fortneyGiantPlanetInterior2010}.

@@

We adopt the system age derived in \citet{cummingsWhiteDwarfInitialFinal2018} using MIST isochrones of \qty{225}{\mega\year}, which implies a ZAMS to TGB age of \qty{99}{\mega\year}. These age determinations are limited both by the precision of the stellar parameters as well as the stellar evolution model. The spread of ages derived by \citet{cummingsWhiteDwarfInitialFinal2018} from different models is $\sim$\qty{10}{\mega\year}. Determining stellar ages is challenging, and our age uncertainty of $\sim$4\% is exceptional compared to those typically obtained by dynamical analysis of young moving groups ($\sim$10\%) or gyrochronology ($\sim$15\%). The values for our adapted and derived parameters are compiled in \Cref{tbl:system}.

@@

By taking data with the field rotator disabled (ADI), the stellar PSF will not appear to rotate but the FOV will appear to rotate. This allows for the effective separation of companion light from the PSF by spatial decorrelating speckles.  After PSF subtraction, we derotated the frames according to their parallactic angle and collapsed the residuals with a variance-weighted sum \citep{bottomNoiseweightedAngularDifferential2017a}, which reduces the pixel-to-pixel noise as the number of frames in the data cube increases.

@@

We determined the best-performing algorithms for each epoch using the contrast curves described in \Cref{sec:analysis}. For the first two epochs, full-frame median subtraction had the best contrast at almost all separations. For the last epoch, annular PCA subtraction with 2 principal components and a rotation threshold of 0.5 FWHM produced the best contrast at close separations (\qtyrange{0.2}{0.4}{\arcsecond}) and had similar performance to other algorithms beyond \ang{;;0.4}. This algorithm was unable to detect a 100 S/N companion injected into the innermost annulus with 5$\sigma$ significance. The contrast for this innermost annulus is therefore not plotted. \Crefrange{fig:residuals}{fig:detection} show the collapsed residual frames from each epoch, along with the Gaussian significance maps and STIM maps.

@@

In this work, we presented high-contrast images of Sirius B in the near-IR. Our 5$\sigma$ sensitivity limits are the best that have been reached for Sirius B so far, reaching \num{20.4} \Lp absolute magnitude at \textgreater\qty{1}{\au}. We consider multiple planetary formation pathways yielding ages between \qtyrange{126}{225}{\mega\year} and explore the effects of enriched metallicity. We translate our sensitivity limits using the ATMO2020 and Sonora Bobcat grids to constrain our mass detection limits to \qtyrange{0.7}{1.1}{\jupitermass} at \textgreater\qty{1}{\au}. Our observations also show how the high precision of the parameters of the Sirius system directly benefits the sensitivity to planets. For example, the $\sim$4\% relative age uncertainty (\Cref{sec:sirius}) translates to a mass uncertainty below \qty{0.1}{\jupitermass}. Despite the high sensitivity of this study, we found no significant evidence for a companion around Sirius B, consistent with previous results.

Although our observations yield no detections, we illustrate the capability of modern high-contrast instrumentation, even without coronagraphy, to reach strong detection limits. Our detection limits benefit directly from the precise stellar characterization of the Sirius system, as well as the proximity and brightness of Sirius B. With laser guide stars \citep[LGS; e.g.,][]{vandamKeckObservatoryLaser2006,baranecRoboAO2FacilityRapid2018}, the limiting magnitude for sufficient AO performances, increases significantly ($m^R\lesssim$ \num{19}). We suspect future work using LGS AO will be capable of studying nearby white dwarf systems at the sub-AU and sub-Jupiter-mass scales \citep{holberg25ParsecLocal2016}. Such observations could significantly improve our theories of planetary formation and stellar evolution beyond the main sequence.

@@

All the code used for pre-processing data, reducing data, and generating the figures is available under an open-source license in a GitHub repository.\footnote{\url{https://github.com/mileslucas/sirius-b}} This code includes all of the scripts for generating each figure in this manuscript. The pre-processed data cubes and parallactic angles are available on Zenodo under an open-source license.\footnote{\dataset[10.5281/zenodo.5115225]{\doi{10.5281/zenodo.5115225}}}. Inquiries regarding data and code are welcome.

>>>>

In recent decades thousands of exoplanets have been discovered orbiting stars that will eventually leave the stability of the main-sequence (MS) \citep{akesonNASAExoplanetArchive2013}. The eventual fate of planets around these stars is uncertain due to the large expansion, stellar winds, and high irradiation encountered during giant branch evolution \citep{verasPostmainsequencePlanetarySystem2016}. Despite this, evidence from white dwarf pollution \citep{juraExternallyPollutedWhite2007,xuSpitzerObservationsWhite2012}, debris disks \citep{deruyterKeplerianDiscsPostAGB2006,zuckermanAncientPlanetarySystems2010,koesterFrequencyPlanetaryDebris2014}, and substellar companions \citep[e.g.,][]{luhmanDiscoveryCandidateCoolest2011,vanderburgGiantPlanetCandidate2020,blackmanJovianAnalogueOrbiting2021} culminate to suggest planetary systems beyond the MS are more common than previously thought. Discovery and characterization of post-MS planets is essential to study how they transform and how they interact with their hosts during the most critical phases of stellar evolution.

The post-MS lifetimes of intermediate-mass stars (\qtyrange{1}{8}{\solarmass}) comprises a relatively brief period of giant branch evolution before all nuclear fusion ends and the stars become white dwarfs. As the main-sequence star runs out of hydrogen to burn in its core, it expands to hundreds of times its size, engulfing any companions within the stellar radius. When helium fusion ignites, the giant star becomes three to four orders of magnitude brighter, causing stellar winds and strong irradiation. Eventually, the star runs out of fuel and concludes its nuclear burning, becoming a white dwarf. The white dwarf begins cooling, becoming three to four orders of magnitude dimmer than its MS progenitor.

@@

Previous searches for substellar companions around white dwarfs \citep[e.g.,][]{debesAreThereUnstable2002,hoganDODOSurveyII2009,luhmanDiscoveryCandidateCoolest2011,xuExtremeAOSearchGiant2015a} have primarily focused on detecting wide-orbit planets which survived the giant branch evolution of their hosts. Recent evidence and theoretical works \citep[e.g.,][]{xuSpitzerObservationsWhite2012,koesterFrequencyPlanetaryDebris2014,verasPostmainsequencePlanetarySystem2016,vanderburgGiantPlanetCandidate2020}, though, suggest ``forbidden'' regions of planetary evolution are worth investigating for exotic post-MS planets. These planets would provide crucial insight into planetary system evolution and planet-star interactions during giant branch evolution.

Direct imaging is well-suited for finding planets around white dwarfs due to the intrinsic faintness of the host (compared to a MS star) and lack of spectral features for radial velocity studies \citep{burleighImagingPlanetsNearby2002a}. With direct imaging, exoplanets can be detected and characterized independently from the orbital period of the planet, the stellar radius, or stellar spectrum, which makes it powerful compared to broadband spectral-energy distribution analysis or transit photometry methods. High-contrast imaging pushes direct imaging to its limits using large-aperture telescopes, adaptive optics, coronagraphy, low-noise detectors, image-processing, and observational techniques. Despite the power of high-contrast imaging, very few observations have been carried out for white dwarfs (a recent example is \citealt{pathakHighContrastImaging2021}).

Exoplanets are faint enough that in certain cases only a few photons per second reach the detector. Host stars are typically many orders of magnitude brighter than the thermal emission of giant planets ($\sim10^{\text{-}8}$), and planets have very small angular separations from their hosts, which makes it difficult to disentangle the planet-light from the stellar diffraction pattern \citep{traubDirectImagingExoplanets2010}. In addition, atmospheric seeing and instrumental aberrations greatly reduce the sensitivity to exoplanets due to the manifestation of quasi-static perturbations of the instrumental point-spread function (speckles, see \citealp{guyonExtremeAdaptiveOptics2018}). Adaptive optics (AO) corrects a large percentage of the effects of speckles but has decreasing efficacy for dim targets due to the photon-limited nature of modern AO instrumentation. AO is also paramount for enabling coronagraphy, which attenuates on-axis starlight while transmitting off-axis signal, up to some inner working angle.

Typical high-contrast targets are nearby young objects observed in the near-infrared. Younger planets are brighter, thanks to their latent formation heat \citep{fortneyGiantPlanetInterior2010} and observing them in the near-infrared corresponds with their peak blackbody emission. Nearby targets have larger angular separations between potential planets and their hosts, which makes it easier to separate the planet from the star. Nearby targets are also intrinsically brighter, enabling effective AO control. White dwarfs are atypical high-contrast targets due to their age and sparsity, but their relative faintness is compelling for the reduced star-planet flux ratio. In addition, a post-MS planet would be much younger than the system age of a white dwarf and would therefore still retain a large portion of its latent heat, further motivating high-contrast searches of nearby white dwarfs.

In this work we set out to perform high-contrast observations of Sirius B to search for post-MS planets. In the rest of this report, we will introduce the Sirius system as a potential candidate for post-MS planets, along with previous studies of white dwarf Sirius B (\Cref{sec:sirius}). We will detail our 2020 near-infrared observations of Sirius B with Keck/NIRC2, as well as our processing steps and statistical analysis for companion detection (\Crefrange{sec:obs}{sec:analysis}). Lastly, we will discuss our results within the context of Sirius and post-MS systems, as well as future directions for post-MS imaging (\Crefrange{sec:results}{sec:conclusion}).

@@

The Sirius system is the 7th closest to the sun at \qty{2.7}{\parsec}, consisting of Sirius A, a $J=$-1.36 magnitude A1V star, and Sirius B, a DA white dwarf with a \qty{50}{\year} orbit \citep{gaiacollaborationGaiaMission2016,bondSiriusSystemIts2017,gaiacollaborationGaiaEarlyData2021}. As mentioned previously, the proximity and intrinsic faintness of Sirius B (compared to a MS star) make it compelling for direct imaging. Additionally, the young system age ($\sim$\qty{225}{\mega\year}) means any giant planets would still retain much of their latent formation heat, increasing their luminosity in the IR \citep{fortneyGiantPlanetInterior2010}.

@@

We adopt the system age derived in \citet{cummingsWhiteDwarfInitialFinal2018} using MIST isochrones of \qty{225}{\mega\year}, which implies a ZAMS to TGB age of \qty{99}{\mega\year}. These age determinations are limited both by the precision of the stellar parameters as well as the stellar evolution model. The spread of ages derived by \citet{cummingsWhiteDwarfInitialFinal2018} from different models is $\sim$\qty{10}{\mega\year}. Determining stellar ages is challenging, and the age uncertainty of $\sim$4\% is exceptional compared to those typically obtained by dynamical analysis of young moving groups ($\sim$10\%) or gyrochronology ($\sim$15\%). The values for our adapted and derived parameters are compiled in \Cref{tbl:system}.

@@

By taking data with the field rotator disabled (ADI), the stellar PSF will not appear to rotate but the FOV will appear to rotate. This allows for the effective separation of companion light from the PSF by spatially decorrelating speckles.  After PSF subtraction, we derotated the frames according to their parallactic angle and collapsed the residuals with a variance-weighted sum \citep{bottomNoiseweightedAngularDifferential2017a}, which reduces the pixel-to-pixel noise as the number of frames in the data cube increases.

@@

We determined the best-performing algorithms for each epoch using the contrast curves described in \Cref{sec:analysis}. For the first two epochs, full-frame median subtraction had the best contrast at almost all separations. For the last epoch, annular PCA subtraction with 2 principal components and a rotation threshold of 0.5 FWHM produced the best contrast at close separations (\qtyrange{0.2}{0.4}{\arcsecond}) and had similar performance to other algorithms beyond \ang{;;0.4}. When processed with this algorithm, we were unable to retrieve a 100 S/N injected companion with 5$\sigma$ significance in the innermost annulus. The contrast in this region is ill-determined and therefore not plotted. \Crefrange{fig:residuals}{fig:detection} show the collapsed residual frames from each epoch, along with the Gaussian significance maps and STIM maps.

@@

In this work, we presented high-contrast images of Sirius B in the near-IR. Our 5$\sigma$ sensitivity limits are the best that have been reached for Sirius B so far, down to \num{20.4} \Lp absolute magnitude at \textgreater\qty{1}{\au}. We consider multiple planetary formation pathways yielding ages between \qtyrange{126}{225}{\mega\year} and explore the effects of enriched metallicity. We translate our sensitivity limits using the ATMO2020 and Sonora Bobcat grids to constrain our mass detection limits to \qtyrange{0.7}{1.1}{\jupitermass} at \textgreater\qty{1}{\au}. Our observations also show how the high precision of the parameters of the Sirius system directly benefits the sensitivity to planets. For example, the $\sim$4\% relative age uncertainty (\Cref{sec:sirius}) translates to a mass uncertainty below \qty{0.1}{\jupitermass}. Despite the high sensitivity of this study, we found no significant evidence for a companion around Sirius B, consistent with previous results.

Although our observations yield no detections, we illustrate the capability of modern high-contrast instrumentation, even without coronagraphy, to reach strong detection limits. Our detection limits benefit directly from the precise stellar characterization of the Sirius system, as well as the proximity and brightness of Sirius B. With laser guide stars \citep[LGS; e.g.,][]{vandamKeckObservatoryLaser2006,baranecRoboAO2FacilityRapid2018} the limiting magnitude for sufficient AO performance is improved ($m^R\lesssim$ \num{19}). Future extremely large telescopes will benefit from increased collecting areas and smaller inner working angles due to the larger aperture diameters. The Thirty Meter Telescope (TMT), for example, will have LGS AO in its first-generation instrument suite (NFIRAOS), which will significantly improve the faintness limits of its high-contrast instrumentation. We suspect future work using LGS AO on current and next-generation telescopes will be capable of studying nearby white dwarf systems at the sub-AU and sub-Jupiter-mass scales \citep{holberg25ParsecLocal2016}. Such observations could significantly improve our theories of planetary formation and stellar evolution beyond the main sequence.

@@

All the code used for pre-processing and reducing the data is available under an open-source license in a GitHub repository.\footnote{\url{https://github.com/mileslucas/sirius-b}} This code includes all of the scripts for generating each figure in this manuscript. The pre-processed data cubes and parallactic angles are available on Zenodo under an open-source license.\footnote{\dataset[10.5281/zenodo.5115225]{\doi{10.5281/zenodo.5115225}}} Inquiries regarding data and code are welcome.
